---
title: "Profiles"
author: "Elyas Mouhou et al."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=TRUE)
options(readr.show_col_types = FALSE)
```

## Aim

The aim is to compute average fragmentomic profiles for the gene classes corresponding to the neutrophil ontology.

## Preamble

```{r, message=FALSE, warning=FALSE}
library('tidyverse')
library('glue')
library('matrixStats')
library('RcppRoll')
library('data.table')
library('readr')
# library('ggrepel')
# library('ggpubr')
# library('gridExtra')
# library('cowplot')
# theme_set(theme_bw())
set.seed(123)
```

```{r}
venn = function (A, B, count=FALSE) {
  f = if (count) length else identity
  lapply(
    list( A=generics::setdiff(A, B),
          AB=generics::intersect(A, B),
          B=generics::setdiff(B, A)),
    FUN=f)
}
```

```{r}
createq = function (gcl, sel, category, sort, class, n=25, by=50) {
  gl = if (sel == '100_random_ubi') {
    gcl %>%
      filter(geneGrouping == '100_random_ubi') %>%
      pull(geneID)
  } else {
    gcl %>%
      filter(str_detect(ontology, 'CL_0000775_neutrophil') &
               geneGrouping == sel) %>%
      pull(geneID)
  }
  
  gl = sort(gl)
  
  cat(class, '\t', gl[1:5], '\n')
  
  selected = dat %>% filter(geneID %in% gl)
  profiles = selected %>% select(`-5000`:`5000`) %>% as.matrix
  
  cat('\t', dim(selected), '\n')
  
  q = colQuantiles(profiles, probs=c(.2, .5, .8)) %>%
    as.data.frame() %>%
    rownames_to_column(var='position') %>%
    as_tibble %>%
    mutate(position=as.integer(position),
           mean=colMeans(profiles)) %>%
    rename(low=`20%`, median=`50%`, high=`80%`)
  
  q = lapply(q %>% select(where(is.numeric)), FUN=function (x) roll_mean(x, n=n)) %>% as_tibble()
  
  q$category = category
  q$sorting  = sort
  q$class    = class
  
  q[seq(1, nrow(q), by=by), ]
}
```

## Data

### All gene classes

How to define the gene classes

```{r}
files = list.files('../data/91_groupsForPubli/')
names(files) = c('coding_FC', 'coding_P', 'lincRNA_FC', 'lincRNA_P')
files = as.list(files)

colnms = c('index',
           'geneID',
           'fold',
           'mann_whitney_pval',
           'chrom',
           'CAT_geneClass',
           'geneGrouping',
           'ontology')

readfrom = function (fn) {
  fn = glue('../data/91_groupsForPubli/{fn}')
  # cat('READING', fn, '\n')
  dat = read_tsv(fn, skip=1, na='', col_names=colnms)
  # cat(dim(dat), '\n')
  dat
}

gclasses = lapply(files, FUN=readfrom)
```

```{r}
gclasses[[1]] %>%
  filter(str_detect(ontology, 'CL_0000775_neutrophil')) %>%
  group_by(CAT_geneClass, geneGrouping) %>%
  summarise(count=n(), .groups='drop')
```

```{r}
gclasses[[1]] %>%
  group_by(CAT_geneClass, geneGrouping) %>%
  summarise(count=n(), .groups='drop')
```

```{r}
# 
gclasses[[1]] %>%
  filter(geneGrouping == '100_random_ubi' | str_detect(ontology, 'CL_0000775_neutrophil')) %>%
  group_by(CAT_geneClass, geneGrouping) %>%
  summarise(count=n(), .groups='drop')
```

### Reconstitue some of the lists

```{r}
onto = read_table('../data/90_fantom/onto/CL_0000775_neutrophil.tsv.gz')
head(onto)
```

```{r}
onto %>% dplyr::count(geneCategory)
```
The following list corresponds to the (\*) ‘coding gene’ category (\*) using $p$ sorting (\*) PF class:

```{r}
gene_list_A = onto %>% filter(geneCategory == 'coding_mRNA') %>%
  arrange(fold)  %>%
  arrange(mann_whitney_pval) %>%
  filter(fold > 0) %>%
  slice_head(n=100) %>%
  pull(geneID)
```

Let's check it's similar to what comes out of the gene classes loaded above

```{r}
gene_list_B = gclasses[[2]] %>%
  filter(str_detect(ontology, 'CL_0000775_neutrophil') &
           geneGrouping == 'first_100_values_with_pos_FC') %>%
  pull(geneID)
```

```{r}
venn(gene_list_A, gene_list_B, count=TRUE)
```
```{r}
venn(gene_list_A, gene_list_B, count=FALSE)
```


The following list corresponds to the (\*) ‘coding gene’ category (\*) using  sorting (\*) NF class:

```{r}
gene_list_A = onto %>% filter(geneCategory == 'coding_mRNA') %>%
  arrange(mann_whitney_pval) %>%
  filter(fold < 0) %>%
  slice_head(n=100) %>%
  pull(geneID)
```

Let's check it's similar to what comes out of the gene classes loaded above

```{r}
gene_list_B = gclasses[[2]] %>%
  filter(str_detect(ontology, 'CL_0000775_neutrophil') &
           geneGrouping == 'first_100_values_with_neg_FC') %>%
  pull(geneID)
```

```{r}
venn(gene_list_A, gene_list_B, count=TRUE)
```
The following list corresponds to the (\*) ‘coding gene’ category (\*) using FC sorting (\*) PF class:

```{r}
gene_list_A = onto %>% filter(geneCategory == 'coding_mRNA') %>%
  arrange(desc(fold)) %>%
  slice_head(n=100) %>%
  pull(geneID)
```

Let's check it's similar to what comes out of the gene classes loaded above

```{r}
gene_list_B = gclasses[[1]] %>%
  filter(str_detect(ontology, 'CL_0000775_neutrophil') &
           geneGrouping == 'first_100_values') %>%
  pull(geneID)
```

```{r}
venn(gene_list_A, gene_list_B, count=TRUE)
```

The following list corresponds to the (\*) ‘coding gene’ category (\*) using FC sorting (\*) NF class:

```{r}
gene_list_A = onto %>% filter(geneCategory == 'coding_mRNA') %>%
  arrange(fold) %>%
  slice_head(n=100) %>%
  pull(geneID)
```

Let's check it's similar to what comes out of the gene classes loaded above

```{r}
gene_list_B = gclasses[[1]] %>%
  filter(str_detect(ontology, 'CL_0000775_neutrophil') &
           geneGrouping == 'last_100_values') %>%
  pull(geneID)
```

```{r}
venn(gene_list_A, gene_list_B, count=TRUE)
```
There is a discrepancy between the lists that may be due to TSS cluster considerations.

## Read coding genes' profiles

### Coding profiles

```{r}
dir = '../../cnam_cfdna_NOTSYNC/data/02_tss_score'
fn  = 'FD_coding_mRNAscoreTable.tsv'
path = glue('{dir}/{fn}')
dat = read_tsv(path, n_max=3)
```

```{r}
n = nrow(dat)
p = ncol(dat)
dat[, (p - 10):p]
```

```{r}
dat = read_tsv(path)
n = nrow(dat)
p = ncol(dat)
```

There are $n = 20244$ genes and $p = 10006$ variables,
including $10001$ positional variables (named `-5000` to `+5000`),
`geneID`, `chrom`, `strand` and `start`.

The following list corresponds to the (a) ‘coding gene’ category (b) using $p$ sorting (c) PF class. This is because we use `gclasses[[2]]` (a, b) and refer to the ‘first_100_values_with_pos_FC’.

```{r}
q = createq(gclasses[[1]],                  # coding | FC
            'first_100_values_with_pos_FC', # PF
            'coding', 'FC', 'PF')

all = q
#     bind_rows(all, q)

q = createq(gclasses[[1]],                  # coding | FC
            'first_100_values_with_neg_FC', # NF
            'coding', 'FC', 'NF')

all = bind_rows(all, q)

q = createq(gclasses[[1]],                  # coding | FC
            'last_100_values',              # b
            'coding', 'FC', 'b')

all = bind_rows(all, q)

q = createq(gclasses[[1]],                  # coding | FC
            'middle_100_values',            # mid
            'coding', 'FC', 'mid')

all = bind_rows(all, q)

q = createq(gclasses[[1]],                  # coding | FC
            '100_random_ubi',               # u
            'coding', 'FC', 'u')

all = bind_rows(all, q)
```

```{r}
q = createq(gclasses[[2]],                  # coding | P
            'first_100_values_with_pos_FC', # PF
            'coding', 'P', 'PF')

all = bind_rows(all, q) # q

q = createq(gclasses[[2]],                  # coding | P
            'first_100_values_with_neg_FC', # NF
            'coding', 'P', 'NF')

all = bind_rows(all, q)

q = createq(gclasses[[2]],                  # coding | P
            'last_100_values',              # b
            'coding', 'P', 'b')

all = bind_rows(all, q)

q = createq(gclasses[[2]],                  # coding | P
            'middle_100_values',            # mid
            'coding', 'P', 'mid')

all = bind_rows(all, q)

q = createq(gclasses[[2]],                  # coding | P
            '100_random_ubi',               # u
            'coding', 'P', 'u')

all = bind_rows(all, q)
```
```{r}
q = createq(gclasses[[3]],                  # lincRNAs | FC
            '100_random_ubi',               # u
            'lincRNAs', 'FC', 'u')

all = bind_rows(all, q)

q = createq(gclasses[[4]],                  # lincRNAs | P
            '100_random_ubi',               # u
            'lincRNAs', 'P', 'u')

all = bind_rows(all, q)
```

### Non coding profiles

```{r}
dir = '../../cnam_cfdna_NOTSYNC/data/02_tss_score'
fn  = 'FD_lncRNA_intergenicscoreTable.tsv'
path = glue('{dir}/{fn}')
dat = read_tsv(path, n_max=3)
```

```{r}
n = nrow(dat)
p = ncol(dat)
dat[, (p - 10):p]
```

```{r}
dat = read_tsv(path)
n = nrow(dat)
p = ncol(dat)
```

```{r}
q = createq(gclasses[[3]],                  # lincRNAs | FC
            'first_100_values_with_pos_FC', # PF
            'lincRNAs', 'FC', 'PF')

all = bind_rows(all, q) # q

q = createq(gclasses[[3]],                  # lincRNAs | FC
            'first_100_values_with_neg_FC', # NF
            'lincRNAs', 'FC', 'NF')

all = bind_rows(all, q)

q = createq(gclasses[[3]],                  # lincRNAs | FC
            'last_100_values',              # b
            'lincRNAs', 'FC', 'b')

all = bind_rows(all, q)

q = createq(gclasses[[3]],                  # lincRNAs | FC
            'middle_100_values',            # mid
            'lincRNAs', 'FC', 'mid')

all = bind_rows(all, q)
```

```{r}
q = createq(gclasses[[4]],                  # lincRNAs | P
            'first_100_values_with_pos_FC', # PF
            'lincRNAs', 'P', 'PF')

all = bind_rows(all, q) # q

q = createq(gclasses[[4]],                  # lincRNAs | P
            'first_100_values_with_neg_FC', # NF
            'lincRNAs', 'P', 'NF')

all = bind_rows(all, q)

q = createq(gclasses[[4]],                  # lincRNAs | P
            'last_100_values',              # b
            'lincRNAs', 'P', 'b')

all = bind_rows(all, q)

q = createq(gclasses[[4]],                  # lincRNAs | P
            'middle_100_values',            # mid
            'lincRNAs', 'P', 'mid')

all = bind_rows(all, q)
```

```{r}
head(all)
```

### Figure

```{r}
all %>%
  mutate(category_sorting=paste(category, sorting, sep='_'),
         category_sorting=fct_relevel(category_sorting,
                                      c('coding_P', 'coding_FC',
                                        'lincRNAs_P', 'lincRNAs_FC')),
         class=fct_relevel(class, c('PF', 'NF', 'mid', 'b', 'u'))) %>%
  ggplot(aes(position, mean)) +
  geom_ribbon(aes(ymin=low, ymax=high), fill='lightgray') +
  geom_line() +
  facet_grid(class ~ category_sorting,
             labeller=labeller(category_sorting=c(
               coding_P='A  Coding - P sorting',
               coding_FC='B  Coding - FC sorting',
               lincRNAs_P='C  LincRNAs - P sorting',
               lincRNAs_FC='D  LincRNAs - FC sorting'))) +
  theme(strip.background=element_rect(color=NA, fill=NA,
                                      size=1.5, linetype="solid"),
        axis.text.x=element_text(angle=45, hjust=1)) +
  xlab('Position wrt TSS')
```


```{r, fig.asp=1.2}
ggsave(filename='all_profiles.png')
```

